Paleoenvironmental models for Australia and the impact of aridification on blindsnake diversification

### 2023-07-10 manuscript.Rmd
second revisions submitted to JBI

**Running title** Blindsnake speciation slows down with aridification

@@

Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia’s terrestrial reptiles [@byrneBirth2008].Of these, roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus *Anilios*, which currently comprises 47 described species **47** described species [@uetzQuarter2021;@tiatragulAustralia2023]. All but three species of currenty described _Anilios_ are endemic to Australia, _A. torresianus_ and _A. leucoproctus_ also occur in southern parts of New Guinea, while _A. eryncinus_ has only been recorded in New Guinea and the Maluku Islands. 
Blindsnakes are morphologically conserved and previous phylogenetic and morphological studies have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single dispersal event by a mesic-adapted ancestor from Southeast Asia or Indonesia between 39 – 19 Ma [@vidalBlindsnake2010;@marinTracing2013]. 
However, phylogenetic analyses from @tiatragulAustralia2023 indicated that the closest relative to all *Anilios* is *Ramphotyphlops multilineatus*, a species that occur in Western New Guinea, Kai Islands, and Salawati, suggesting the Australian radiation may have a Melanesian origin.
Despite representing a substantial portion of snake diversity, we know little about the macroevolutionary dynamics of blindsnakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic environmental changes during the Cenozoic because they are geographically restricted and can be found in every biome on mainland Australia.

Advances in phylogenetic comparative methods and paleoenvironmental reconstructions allow us to better understand regional diversification dynamics. The influence of temperature on diversification rates has been empirically tested with birth-death models [@condamineMacroevolutionary2013;@kergoatOpposite2018;@condamineClimate2019;@rollandContribution2019]. Global paleotemperature data for these analyses was first presented in @condamineMacroevolutionary2013, whereby relative proportions of different oxygen isotopes in samples of benthic foraminifer shells [@zachosEarly2008] over geological time was converted to absolute temperature using a formula from @epsteinRevised1953. The relationship between diversification rates and aridity, however, have been attributed to some proxies such as the timing of aridification or global temperature rather than quantifying the extent of regional aridity itself [@dornStrange2014;@deepakAridification2018;@catulloAridification2014;@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014;@owenHow2017;@crispHow2013;@toussaintUnveiling2015].

To our knowledge no study has assessed aridity as a continuous variable in an environment-dependent birth-death model framework, probably due to the difficulty of quantifying paleoaridity. However, recent developments in spatially-explicit paleoenvironmental reconstructions from the lithological and fossil record as well as global circulation models (GCMs) now enables us to estimate paleotemperature and paleoaridity data specific to Australia in the time span of interest. The use of these data sets allows us to conduct empirical analyses of the impacts of regionalised environmental changes on diversification rates over time for the first time.

In this study, we aimed to infer the historical biogeography of the Australian blindsnake radiation and investigate the impact of paleoenvironmental changes, specifically changes in temperature and aridity associated with Neogene aridification, on diversification rates within this group. To do this, we fitted several birth-death models with reconstructed paleotemperature and aridity, and estimated diversification rates under these models. Additionally, we compared diversification rates between arid-adapted lineages versus mesic-adapted lineages to investigate if lineages that inhabit the arid biome have undergone faster or slower diversification.

@@

With 47 recognised species, the mostly Australian-endemic blindsnake genus *Anilios* is currently the most diverse genus of Scolecophidia in the world. In this study we estimated that the ancestral species in Australia occupied the tropical grassland biome, most likely as a result of invasion from other tropical mesic areas including New Guinea, Bismarck Archipelago, and the Lesser Sundas. Within a few million years, this ancestral lineage split with one group staying in the tropical grassland biome while the other group transitioned into the arid biome. Subsequent biome transitions in these two groups resulted in the distribution of biomes we observed across the phylogeny today. We also found that the paleoaridity index provides the best fit to the diversification dynamics observed in our time-calibrated phylogeny. Although aridification has been linked to elevated diversification rates in other plants and animals [@brennanMass2017;@raboskyAnalysis2014;@hernandez-hernandezAridification2014;@jhaPhylogeny2021], our results suggest that aridification decreased diversification rates for Australian *Anilios* blindsnakes.

@@

Our best-fitting diversification model show that speciation is negatively correlated with aridity (Fig. \@ref(fig:fig-RPANDA)c). Our reconstructed paleoaridity index data indicate that aridity increased from ~25 Ma following the opening of the Southern Ocean and the onset of the Antartic circumpolar current [@byrneBirth2008]. A particularly sharp increase from 10 Ma followed from the "Hill Gap" period of cooling [@hillHistory1994], and intensification from 5 Ma onwards following the Pliocene transition to a contemporary arid Australian climate [@byrneBirth2008]. Out of the 38 lineages we considered, 5/35 divergence events occurred in the last 10 Ma when the continent intensively aridified. In contrast, 31/35 divergence event occurred between 25 – 10 Ma when the continent was getting warmer but not entirely arid. Indeed 12/16 lineages that have current distributions in the arid biome diverged before 10 Ma. This pattern is consistent with other studies on vertebrates that have shown that the divergence time for some arid-adapted taxa pre-dated the intense aridification in the last 5 Ma [@byrneBirth2008;@catulloAridification2014].

The observed decrease in diversification rate with increasing aridification may be linked to decreasing speciation rates or increasing extinction rates. One hypothesis is that mesic-adapted ancestral blindsnakes may have been widespread in the past, but populations later became extinct due to extreme aridity in the last 5 Ma [@fujiokaAustralian2009;@byrneDecline2011].
An extensive population genetics study of blindsnake by @marinHidden2013 proposed that during widespread aridification, mesic-adapted lineages went extinct or were confined to mesic refugia, such as the Pilbara and Kimberley, where they became pre-adapted to seasonal aridity, which enable subsequent invasion into the truly arid environments.
Additionally, long naked branches in our current phylogeny, particularly in Group 1, suggests past extinction events or intraspecific diversity not yet recognised.
Our best-fitting models of diversification support constant extinction rates with decreasing rates of speciation. 
However, the lack of *Anilios* fossils makes extinction rates difficult to estimate [@loucaWhy2021;@raboskyExtinction2010]. An alternative explanation for 
the negative relationship between aridification and diversification is decreasing speciation rates through time. This could be driven by the indirect effect of aridification on population size [@goldieEvolutionary2010]. More arid sites have lower primary productivity and may support smaller population sizes. Under the integrated evolutionary speed hypothesis [@gillmanPatterns2013;@rohdeLatitudinal1992], smaller population sizes are predicted to have slower rates of molecular evolution and subsequently lower rates of speciation. Lower rates of molecular evolution have been found in Australian arid zone plants compared to mesic zone plants [@goldieEvolutionary2010]. 
Yet, here we found no difference between arid zone and mesic zone speciation rates in blindsnakes which is inconsistent with this hypothesis. 
Finally, another potential explanation for the negative correlation between diversification and aridity is the limited sampling of Australian blindsnakes, particularly in major sand deserts [@pepperLife2021] or even in areas of high cryptic diversity such as the Kimberley [@ellisTyphlopid2019]. The current phylogeny is missing 12 described species [@tiatragulAustralia2023], such as the arid-adapted _A. fossor_ and _A. minimus_, and the true diversity has been estimated to exceed 56 species [@marinHidden2013;@marinTracing2013;@ellisTyphlopid2019]. It is possible that unrecognised species may be hidden under nominal species and that these missing lineages have diversified more recently when aridity is high. If all suspected cryptic OTUs were treated as recognised species, we could see a different relationship with aridification.

## Diversification rates in arid vs mesic biomes
Several studies have shown that arid-adapted species originated either from a single ancestor present in Australia before the Miocene, or through multiple independent expansion into the arid zone [@byrneBirth2008]. Our ancestral biome analyses show that several mesic ancestors of _Anilios_ blindsnakes have diverged into the arid zone evident by multiple arid species throughout the phylogeny. This pattern is consistent with several other groups of reptiles that started to diversify in the Miocene including monitor lizards [@brennanPhylogenomics2021;@pavon-vazquezCompetition2022], _Sphenomorphus_ and _Eugongylus_ skinks [@skinnerLygosomine2011], elapid snakes [@sandersMolecular2008], pythons [@esquerrePhylogenomics2020;@esquerreEvolutionary2022] and several geckos [@ashmanDiversification2018;@fujitaDiversification2010;@oliverIndependent2014]. For these terrestrial reptiles, increased rates of speciation for arid-adapted taxa have been attributed to species filling niches in the saxicolous and xeric habitats that emerged with the contraction of mesic forest throughout the Miocene [@sandersMolecular2008;@hugallPhylogeny2008]. However, results from our weight-averaged SSE analyses on the Australian blindsnake phylogeny indicated that speciation, extinction, and turnover rates did not differ between lineages in arid and mesic biomes. This lack of difference between biomes could be due in part to unrecognised species from the often under sampled arid zone [@pepperLife2021]

@@

Our BioGeoBEARS ancestral biome estimation is limited by the areas (e.g. in this case biomes) to which we can assign our extant species. By following the WWF Bioregions maps, we had to assign species that occur in extremely different habitats in the same biome. For example, *A. longissimus*, a species that has only been collected from 30 m underground on Barrow Island located 50 km the northwest off the Pilbara coast of Western Australia, shares the same biome with *A. centralis* from the arid interior of the continent. Current methods do not allow us to account for heterogeneity of the biome even though we know the arid biome consists of everything from flat, seemingly featureless, desert in the interior to topologically complex mountain ranges on the fringes [@pepperLife2021;@pepperGenetic2011]. However, modern software under development such as PhyBEARS may be able to account for more complex geological processes such as the movement of continents. Finally, (mass) extinction likely played important roles in the pattern of diversity we observed today [@oliverPhylogenetic2017]. Out of all our comparative analyses, only the SSE and some RPANDA models estimated extinction rates.

>>>>

**Running title** Blindsnake speciation slows down with increased aridification

@@

Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia’s terrestrial reptiles [@byrneBirth2008].Of these, roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus *Anilios*, which currently comprises 47 described species **47** described species [@uetzQuarter2021;@tiatragulAustralia2023]. All but three species of formally described _Anilios_ are endemic to Australia, _A. torresianus_ and _A. leucoproctus_ also occur in southern parts of New Guinea, while _A. eryncinus_ has only been recorded in New Guinea and the Maluku Islands. 
Blindsnakes are morphologically conserved and previous phylogenetic and morphological studies have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single dispersal event by a mesic-adapted ancestor from Southeast Asia or Indonesia between 39 – 19 Ma [@vidalBlindsnake2010;@marinTracing2013]. 
However, phylogenetic analyses from @tiatragulAustralia2023 indicated that the closest relative to all *Anilios* is *Ramphotyphlops multilineatus*, a species that occur in Western New Guinea, Kai Islands, and Salawati, suggesting the Australian radiation may have a Melanesian origin.
Despite representing a substantial portion of snake diversity, we know little about the macroevolutionary dynamics of blindsnakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic environmental changes during the Cenozoic because they are geographically restricted and can be found in every biome on mainland Australia.

Advances in phylogenetic comparative methods and paleoenvironmental reconstructions allow us to better understand regional diversification dynamics. The influence of temperature on diversification rates has been empirically tested with birth-death models [@condamineMacroevolutionary2013;@kergoatOpposite2018;@condamineClimate2019;@rollandContribution2019]. Global paleotemperature data for these analyses was first presented in @condamineMacroevolutionary2013, whereby relative proportions of different oxygen isotopes in samples of benthic foraminifer shells [@zachosEarly2008] over geological time was converted to absolute temperature using a formula from @epsteinRevised1953. The relationship between diversification rates and aridity, however, have been attributed to some proxies such as the timing of aridification or global temperature rather than quantifying the extent of regional aridity itself [@dornStrange2014;@deepakAridification2018;@catulloAridification2014;@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014;@owenHow2017;@crispHow2013;@toussaintUnveiling2015]. To our knowledge no study has assessed aridity as a continuous variable in an environment-dependent birth-death model framework, probably due to the difficulty of quantifying paleoaridity. However, recent developments in spatially-explicit paleoenvironmental reconstructions from the lithological and fossil record as well as global circulation models (GCMs) now enables us to estimate paleotemperature and paleoaridity data specific to Australia in the time span of interest. The use of these data sets allows us to conduct empirical analyses of the impacts of regionalised environmental changes on diversification rates over time for the first time.

In this study, we aimed to infer the historical biogeography of the Australian blindsnake radiation and investigate the impact of paleoenvironmental changes, specifically changes in temperature and aridity associated with Neogene aridification, on diversification rates within this group. To do this, we fitted several birth-death models with reconstructed paleotemperature and aridity, and estimated diversification rates under these models. Additionally, we compared diversification rates between arid-adapted lineages versus mesic-adapted lineages to investigate if lineages that inhabit the arid biome have undergone faster or slower diversification. We predicted that arid-adapted lineages will have higher diversification rates compared to mesic-adapted lineages. This expectation is based on the hypothesis that aridification can cause  fragmentation of continuous habitats, leading to isolated populations in distinct microhabitats and refugia in arid areas, which may lead to greater species level divergence [@brennanMiocene2018]. Additionally, the ecological opportunities or selective pressures resulting from aridification may drive the evolution of specialised traits in mesic-adapted ancestral species, enabling them to exploit previously unoccupied niches and promote speciation.

@@

With 47 recognised species, the mostly Australian-endemic blindsnake genus *Anilios* is currently the most diverse genus of Scolecophidia in the world. In this study we estimated that the ancestral species in Australia occupied the tropical grassland biome, most likely as a result of dispersal from other tropical mesic areas including New Guinea, Bismarck Archipelago, and the Lesser Sundas. Within a few million years, this ancestral lineage split with one group staying in the tropical grassland biome while the other group transitioned into the arid biome. Subsequent biome transitions in these two groups resulted in the distribution of biomes we observed across the phylogeny today. We also found that the paleoaridity index provides the best fit to the diversification dynamics observed in our time-calibrated phylogeny. Although aridification has been linked to elevated diversification rates in other plants and animals [@brennanMass2017;@raboskyAnalysis2014;@hernandez-hernandezAridification2014;@jhaPhylogeny2021], our results suggest that aridification decreased diversification rates for Australian *Anilios* blindsnakes.

@@

Our best-fitting diversification model show that speciation is negatively correlated with aridity (Fig. \@ref(fig:fig-RPANDA)c). Our reconstructed paleoaridity index data indicate that aridity increased from ~25 Ma following the opening of the Southern Ocean and the onset of the Antarctic circumpolar current [@byrneBirth2008]. A particularly sharp increase from 10 Ma followed from the "Hill Gap" period of cooling [@hillHistory1994], and intensification from 5 Ma onwards following the Pliocene transition to a contemporary arid Australian climate [@byrneBirth2008]. Out of the 38 lineages we considered, 5/35 divergence events occurred in the last 10 Ma when the continent intensively aridified. In contrast, 31/35 divergence event occurred between 25 – 10 Ma when the continent was getting warmer but not entirely arid. Indeed 12/16 lineages that have current distributions in the arid biome diverged before 10 Ma. This pattern is consistent with other studies on vertebrates that have shown that the divergence time for some arid-adapted taxa pre-dated the intense aridification in the last 5 Ma [@byrneBirth2008;@catulloAridification2014].

The observed decrease in diversification rate with increasing aridification may be linked to decreasing speciation rates or increasing extinction rates. One hypothesis is that mesic-adapted ancestral blindsnakes may have been widespread in the past, but populations later became extinct due to extreme aridity in the last 5 Ma [@fujiokaAustralian2009;@byrneDecline2011].
An extensive population genetics study of blindsnake by @marinHidden2013 proposed that during widespread aridification, mesic-adapted lineages went extinct or were confined to mesic refugia, such as the Pilbara and Kimberley, where they became pre-adapted to seasonal aridity, which enable subsequent invasion into the more arid environments.
Additionally, long naked branches in our current phylogeny, particularly in Group 1, suggests past extinction events or intraspecific diversity not yet recognised.
Our best-fitting models of diversification support constant extinction rates with decreasing rates of speciation. 
However, the lack of *Anilios* fossils makes extinction rates difficult to estimate [@loucaWhy2021;@raboskyExtinction2010]. An alternative explanation for 
the negative relationship between aridification and diversification is decreasing speciation rates through time. This could be driven by the indirect effect of aridification on population size [@goldieEvolutionary2010]. More arid sites have lower primary productivity and may support smaller population sizes. Under the integrated evolutionary speed hypothesis [@gillmanPatterns2013;@rohdeLatitudinal1992], smaller population sizes are predicted to have slower rates of molecular evolution and subsequently lower rates of speciation. Lower rates of molecular evolution have been found in Australian arid zone plants compared to mesic zone plants [@goldieEvolutionary2010]. 
Yet, here we found no difference between arid biome and mesic biome speciation rates in blindsnakes which is inconsistent with this hypothesis. 
Finally, another potential explanation for the negative correlation between diversification and aridity is the limited sampling of Australian blindsnakes, particularly in major sand deserts [@pepperLife2021] or even in areas of high cryptic diversity such as the Kimberley [@ellisTyphlopid2019]. The current phylogeny is missing 12 described species [@tiatragulAustralia2023], such as the arid-adapted _A. fossor_ and _A. minimus_, and the true diversity has been estimated to exceed 56 species [@marinHidden2013;@marinTracing2013;@ellisTyphlopid2019]. It is possible that unrecognised species may be hidden under nominal species and that these missing lineages have diversified more recently when aridity is high. If all suspected cryptic OTUs were treated as recognised species, we could see a different relationship with aridification.

## Diversification rates in arid vs mesic biomes
Several studies have shown that arid-adapted species originated either from a single ancestor present in Australia before the Miocene, or through multiple independent expansion into the arid zone [@byrneBirth2008]. Our ancestral biome analyses show that several mesic ancestors of _Anilios_ blindsnakes have diverged into the arid zone evident by multiple arid species throughout the phylogeny. This pattern is consistent with several other groups of reptiles that started to diversify in the Miocene including monitor lizards [@brennanPhylogenomics2021;@pavon-vazquezCompetition2022], _Sphenomorphus_ and _Eugongylus_ skinks [@skinnerLygosomine2011], elapid snakes [@sandersMolecular2008], pythons [@esquerrePhylogenomics2020;@esquerreEvolutionary2022] and several geckos [@ashmanDiversification2018;@fujitaDiversification2010;@oliverIndependent2014].

For these terrestrial reptiles, increased rates of speciation for arid-adapted taxa have been attributed to species filling niches in the saxicolous and xeric habitats that emerged with the contraction of mesic forest throughout the Miocene [@sandersMolecular2008;@hugallPhylogeny2008]. Contrary to our predictions, results from our analyses on the Australian blindsnake phylogeny revealed no differences in speciation, extinction, and turnover rates between arid and mesic lineages. This suggests that the influence of aridity on diversification rates is more complex and does not follow the pattern we expected if species are filling new niches. Several factors may explain the lack of difference in diversification rates between biomes. One possible biological reason is that ancestral mesic-adapted blindsnakes already possessed specialised traits suitable for both arid and mesic subterranean environments, which reduced the potential for divergent diversification rates. A possible lack of differentiation in prey and availability of these prey items, particularly ant and termite larvae and pupae, in both biomes could also contribute to comparable diversification rates observed. Additionally, all species of blindsnakes exclusively inhabit subterranean environments during the day, which shield them against surface climatic variations such as extreme heat in the arid zone [@ehmannFamily1993]. While other reptiles may seek shade cover, many species must also physiologically adapt to extreme temperatures. Finally, incomplete sampling may obscure the true diversification patterns, particularly because there may be a larger proportion of unrecognised species from the arid zone[@pepperLife2021].

@@

Our BioGeoBEARS ancestral biome estimation is limited by the areas (e.g. in this case biomes) to which we can assign our extant species. By following the WWF Bioregions maps, we had to assign species that occur in extremely different habitats in the same biome. For example, *A. longissimus*, a species that has only been collected from 30 m underground on Barrow Island located 50 km the northwest off the Pilbara coast of Western Australia, shares the same biome with *A. centralis* from the arid interior of the continent. Current methods do not allow us to account for heterogeneity of the biome even though we know the arid biome consists of everything from flat, seemingly featureless, desert in the interior to topologically complex mountain ranges on the fringes [@pepperLife2021;@pepperGenetic2011]. However, modern software under development such as PhyBEARS may be able to account for more complex geological processes such as the movement of continents. For our SSE analyses, it is also noteworthy that the lineages we designated as “mesic-adapted” in reality may occupy a variety of biomes with varying degree of aridity from tropical rainforests to tropical savannah. Finally, (mass) extinction likely played important roles in the pattern of diversity we observed today [@oliverPhylogenetic2017]. Out of all our comparative analyses, only the SSE and some RPANDA models estimated extinction rates.
